---
title: "A report on the SILC data for Belgium"
author: "Anne Vinkel Hansen"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
par(mar = c(0,0,0,0))
```

This report is generated in Rstudio using markdown. To generate a Word document 
from this .Rmd file, press the "Knit" icon and choose "Knit to Word".

## The data

We use `read.delim` to read in data here, but recall `read_sav` in the `haven` 
package, and `read_xlsx` in the `readxl`` package. 

```{r, include=F}
### Reading in data

library(lubridate) # For handling dates 

silc_p <- read.delim("BE_2013p_EUSILC.csv", sep=";")
```

The data is the SILC data for Belgium and is stored at this path: 

`r getwd()`. 

It contains data on Statistics on Income and Living Conditions for the year(s) 
`r min(silc_p$PB110)`-`r `max(silc_p$PB110)`. 

## Some tables

We shall tabulate data by sex, marital status and year of birth. 

```{r, echo = F, message= F, warning =F}

# First, we want to group year of birth into intervals

silc_p$YEAR_OF_BIRTH <- cut( silc_p$PB140
                           , breaks = seq( from = 1930, to = 2010, by=20)
                           , right = F
                           , labels = c("1930-49", "1950-69", "1970-89","1990-2009"))

# We want a more informative format on Sex and marital status

silc_p$SEX <- factor( silc_p$PB150, levels = c(1,2), labels = c("Male", "Female"))

silc_p$MARITAL_STATUS = factor( silc_p$PB190, 
                                levels = 1:5,
                                labels = c( "Never married","Married","Separated","Widowed","Divorced")
)

```
Basic tables are functional but somewhat lacking. They can be made prettier with 
the `knitr` package. 

```{r, echo = F, message= F}
library(knitr)
library(dplyr)

table( silc_p$YEAR_OF_BIRTH, silc_p$SEX) %>% kable()
```

For a more powerful, if syntactically heavy approach, the `tables` package is an 
option. See `vignette(tables)`

```{r, echo = F, message= F}
library(tables)
library(kableExtra)

# Syntax that makes a table
tab <- tabular( 
    1 + YEAR_OF_BIRTH + MARITAL_STATUS ~ 
      SEX*( (N=1) + Percent("col")*Format(digits=1))
  , data = silc_p
  )

tab
```

We may want to do some descriptive plots, say of year of birth by sex

```{r, echo=FALSE, message=FALSE}
library(ggplot2)

ggplot( data = silc_p
      , aes( x = PB140)) + 
  geom_histogram( aes( y =..density.. )) + 
  geom_density( col = "magenta") +
  facet_wrap( ~ SEX) + 
  xlab( "Year of birth" )
```



