---
title: 'Summary and workshop'
author: "Anne Vinkel"
date: "Statistical Programming with R"
output:
  ioslides_presentation:
    fig_height: 5
    fig_width: 8
    logo: logo_en.png
    transition: faster
    smaller: true
    html_document:
    self_contained: false
    lib_dir: libs
  beamer_presentation: default
---

## This lecture

- Summary 

- Workshop introduction

# Using R

## Help 

- Google is your friend
- As is the help file. 
  - If you know the name of the function that performs an operation, e.g. `anova()`, then you just type `?anova` or `help(anova)` in the console. 
  - If you do not know the name of the function: type `??` followed by your search criterion. For example `??anova` returns a list of all help pages that contain the word 'anova'


## The basics

- Write commands directly in the console
- Or write code in the editor and submit with Ctrl + Enter
- Assign values to objects with `<-`

```{r, results = "hide"}
a <- c(1, 8, 42, pi, 2^3, 1)
b <- 0:5
c <- rep( c("Hi", "there"), 3)
D <- as.data.frame(cbind(a,b,c))
```

## Objects
Vectors and character vectors
```{r}
c(1,2,3,4,5)
1:5
as.character(1:5)
```

## Data frames
Data sets for R

```{r}
D <- as.data.frame(cbind(a,b,c))
D
```
Alternatively: `tibbles` or `data.tables` 

## Subsetting 

```{r}
D[1,3]
D[1,]
D[,2]
D$a
```


## Pipes and the tidyverse

The tidyverse is a collection of packages that "share an underlying design philosophy, grammar, and data structures". They make for easier data handling and visualization. 

The pipe %>% allows for chained method calls that make code much more readable
```{r, results="hide", message = F}
library(dplyr)
starwars %>%
  subset( species == "Human" ) %>%
  group_by( homeworld ) %>% 
  summarise( n = n(), mean.height = mean(height))
```

## Plotting {.smaller}
Plots in base R are fast and easy:
```{r}
plot(starwars$height, starwars$mass)
hist(starwars$height)
```

## ggplot2
```{r, warning=F, message=F}
library(ggplot2)
starwars %>% ggplot( aes( x = height)) + geom_histogram()
```

# Working in `R`

## Working code-centrically

- A single file of code, or a series of code files stored in the same place produce all your results
- Separates data from results
- Can be run non-interactively
- Allows for reproducibility
- Not unique to R


## Working code-centrically

- Work in projects to make your code more self-contained
- Use comments, spacing and indentation to make your code readable

```{r, eval = F}
# Look at this -----------------------------------------------------------------
str(mammalsleep)
summary(mammalsleep)
mammalsleep[mammalsleep$species%in%c("Echidna","Lesser short-tailed shrew","Musk shrew"),]


# Compared with this -----------------------------------------------------------

# Structure of the "mammalsleep" data in the mice package
str(mammalsleep)

# Data summary
summary(mammalsleep)

# Pick out three specific rows
mammalsleep[ mammalsleep$species %in% 
               c("Echidna", "Lesser short-tailed shrew", "Musk shrew") , ]
```


## The R workflow

- **Three separate things:** The data, the results and the code producing the results
- The code:
   - Reads in data
   - Performs checks, transformations and analyses
   - Creates output
      - Possibly in the form of Rmarkdown reports
- Place the code, the data and the results in sensible places on your computer


## Now what?

- **Now** we do a workshop where you work on geostat data
  - Generate relevant summaries, tables etc
  - I have an example (short and slapdash) report working on SILC data
  
- **After the course**
  - The *further studies* tab on the webpage
  - Try using this on your own data
  - Feel free to write me with questions (aih@dst.dk)
  
# Workshop