---
title: 'Data manipulation'
author: "Anne Vinkel, Gerko Vink & Erik-Jan van Kesteren"
date: "Statistical Programming with R"
output:
  ioslides_presentation:
    fig_height: 5
    fig_width: 8
    logo: logo.png
    transition: faster
    smaller: true
    html_document:
    self_contained: false
    lib_dir: libs
  beamer_presentation: default
---

## This lecture

- Data manipulation

- Basic analysis (correlation & t-test)


## New packages we use
```{r warning=FALSE, message=FALSE}
library(MASS)     # for the cats data
library(haven)    # in/exporting data
library(magrittr) # for pipes
library(dplyr)    # data manipulation
```

# Data manipulation

## The cats data

```{r}
head(cats)
```


```{r}
str(cats)
```

## How to get only Female cats?

```{r}
fem.cats <- cats[cats$Sex == "F", ]
dim(fem.cats)
head(fem.cats)
```

## How to get only *heavy* cats?
```{r}
heavy.cats <- cats[cats$Bwt > 3, ]
dim(heavy.cats)
head(heavy.cats)
```

## How to get only *heavy* cats?
```{r}
heavy.cats <- subset(cats, Bwt > 3)
dim(heavy.cats)
head(heavy.cats)
```

<!-- ## another way: `dplyr` -->
<!-- ```{r} -->
<!-- filter(cats, Bwt > 2, Bwt < 2.2, Sex == "F") -->
<!-- ``` -->


## Working with factors
```{r}
class(cats$Sex)
levels(cats$Sex)
```

## Working with factors
```{r}
levels(cats$Sex) <- c("Female", "Male")
table(cats$Sex)
head(cats)
```

## Sorting 

```{r}
sorted.cats <- cats[order(cats$Bwt), ]
head(sorted.cats)
```


## Combining matrices or dataframes

```{r}
cats.wider <- cbind( cats, 1:144)
head(cats.wider)
```

## Combining matrices or dataframes
```{r}
rbind(cats[1:3, ], cats[1:5, ])
```


# Basic analysis

## Correlation

```{r fig.height=5, fig.width=5, dev.args = list(bg = 'transparent'), fig.align='center'}
plot(cats$Bwt, cats$Hwt)
```

## Correlation
Let's compute the correlation with the `cor` function. According to the help file (`?cor`), we can use just `cats[,-1]` as input

```{r}
cor(cats[, -1])
```
With `[, -1]` we exclude the first column

## Correlation

```{r}
cor.test(cats$Bwt, cats$Hwt)
```


# Pipes

## Pipes

Pipes are a very useful tool from the `magrittr` package that allow you to chain a 
sequence of operations in a readable and intuitive manner. 

They're also an entirely new syntax, so we will not focus on them in this course. 
Links on the course webpage if you want to learn more. 

## This is a pipe:

```{r message=FALSE}
boys <- 
  read_sav("boys.sav") %>%
  head()
```

It effectively replaces `head(read_sav("boys.sav"))`.


## Why are pipes useful?
Let's assume that we want to load data, change a variable, filter cases and select columns. Without a pipe, this would look like
```{r}
boys  <- read_sav("boys.sav")
boys2 <- transform(boys, hgt = hgt / 100)
boys3 <- filter(boys2, age > 15)
boys4 <- subset(boys3, select = c(hgt, wgt, bmi))
```

With the pipe:
```{r}
boys <-
  read_sav("boys.sav") %>%
  transform(hgt = hgt/100) %>%
  filter(age > 15) %>%
  subset(select = c(hgt, wgt, bmi))
```

Benefit: a single object in memory that is easy to interpret


## With pipes
Your code becomes more readable:

- data operations are structured from left-to-right and not from in-to-out
- nested function calls are avoided
- local variables and copied objects are avoided
- easy to add steps in the sequence

## What do pipes do:

- `f(x)` becomes `x %>% f()`
```{r}
rnorm(10) %>% mean()
```
- `f(x, y)` becomes `x %>% f(y)` 
```{r}
boys %>% cor(use = "pairwise.complete.obs")
```
- `h(g(f(x)))` becomes `x %>% f %>% g %>% h` 
```{r}
boys %>% subset(select = wgt) %>% na.omit() %>% max()
```

## Useful: outlier filtering
```{r}
nrow(cats)

cats.outl <- 
  cats %>% 
  filter(Hwt < mean(Hwt) + 3 * sd(Hwt), 
         Hwt > mean(Hwt) - 3 * sd(Hwt))

nrow(cats.outl)

cats %>%  
  filter(Hwt > mean(Hwt) + 3 * sd(Hwt))
```

# Practical 