---
title: "Loops and functions"
author: "Peter Stoltze and Anne Vinkel Hansen"
date: "Statistical Programming with R"
output:
  # ioslides_presentation:
  #   css: css/style.css
  #   logo: css/logo_en.png
  #   transition: 0
  html_document:
    # css: style_ho.css
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false

---

## This lecture will introduce

- If-statements
- For-loops
- The apply-family
- How to write your own functions

##

### New control flow constructs
- `if(cond)` `expr`
- `if(cond)` `expr.1`  `else`  `expr.2`
- `for(var in seq)` `expr`

### New functions
- `rev()`: reverse version of an argument (backwards)
- `apply()`: apply a function to margins of a matrix
- `sapply()`: apply a function to elements of a list, `vector` or `matrix` return
- `lapply()`: apply a function to elements of a list, `list` return
- `print()`: print an object to the console
- `cat`: outputs an object with less conversion than `print()`

##

- We load the `MASS` packages to have access to the infamous `cats` data set :-)
```{r echo=TRUE, warning=FALSE, message=FALSE}
library(MASS)
summary(cats)
```

# Conditionals and loops

## If-statements

- Often we want to run some code only if some condition is true, e.g.

```{r}
a <- 2
if (a > 5){
  print("a is larger than 5")
}
```

```{r}
a <- 8
if (a > 5){
  print("a is larger than 5.")
}
```

## If-else-statements

- We can also specify an alternative action to be run if the condition is **not** true
```{r}
a <- 2
if (a > 5){
  print("a is larger than 5")
} else {
  print("a is not larger than 5")
}
```

##

```{r}
a <- 8
if (a > 5){
  print("a is larger than 5")
} else {
  print("a is not larger than 5")
}
```

## For-loops

- Loops are used when we want to perform some repetitive calculations 
- It is often tedious, or even impossible, to write this repetition out completely 
- Consider a simple task as printing the numbers from 1 to 6 one-by-one 

## {.small}
```{r}
print(1)
print(2)
print(3)
```

## {.small}
```{r}
print(4)
print(5)
print(6)
```

## 
- A for-loop allows us to automate this!
```{r}
for (i in 1:6){
  print(i)
}
```

##
- The idiomatic choice of counter is `i` but it could be anything
```{r}
for (some.var.name in 1:6){
  print(some.var.name)
}
```

##

```{r}
for (i in 1:6){
  print(i < 5)
}
```

##
- List all out of the 144 cats weighting more than 3.5 kg
```{r}
for (i in 1:nrow(cats)){
  if (cats$Bwt[i] > 3.5){
    cat("Cat no.", i, "weighs more than 3.5 kg, namely", cats$Bwt[i], "kg\n")
  }
}
```

##
- List the class of the different columns in the `cats` data
```{r}
for (i in 1:ncol(cats)){
  print(class(cats[, i]))
}
```


##
- Nested for-loops are quite common
```{r}
for (i in 1:3){
  for (j in 1:2){
    print(paste(i, "x", j, "=", i*j))
  }
}
```

##
- Spot the difference
```{r}
for (i in 1:3){
  for (j in 1:2){
    print(paste0(i, "x", j, "=", i*j))
  }
}
```


## Looping over lists or vectors
- As default behaviour the for-loop traverses all elements
```{r}
my.list <- list(A = c(4, 2, 1:3), B = "Hello.", C = TRUE)

for (list.item in my.list){
  cat("One element is", list.item, "\n")
}
```

##  {.smaller}
- But if we want to change the item, we have to be able to access it, so we are better off using `1:length(list)`
```{r}
MyList <- list(A = c(2, 1:3), B = "Hello", C = c(TRUE, TRUE, FALSE))

for (i in 1:length(MyList)){
  MyList[[i]] <- rev(MyList[[i]])
}
MyList
```

# The `apply()` family

## The `apply()` family

- The `apply` family is a group of very useful functions that allow you to easily execute a function of your choice across an objects, such as a `list`, a `data.frame`, or a `matrix`
- We will look at three members of the family
    - `apply`
    - `sapply`
    - `lapply`

## `apply()`
- `apply` is used for matrices and dataframes
- It can take a function that takes a vector as input and *apply* it to either a *row* or a *column* 
- The usage is `apply(X, MARGIN, FUN, ...)` which means that the first three *arguments* are fixed
    - `X` is the data
    - `MARGIN` is the *direction* in which to apply the function; 1 means rows while 2 means columns
    - `FUN` is the *function* you want to apply

## {.build}
- Calculate the mean for all columns in the cats data sans 1 (which is sex) 

```{r}
apply(cats[, -1], MARGIN = 2, mean)
```

- Of course there is a short cut for this, but the former is more general
```{r}
colMeans(cats[, -1])
```


##
- Create a 4-by-6 matrix consisting of random draws from the standard normal distribution
```{r} 
rand.mat <- matrix(rnorm(24), nrow = 4, ncol = 6)
rand.mat
```

##
```{r}
apply(rand.mat, MARGIN = 1, FUN = max)
apply(rand.mat, MARGIN = 2, FUN = max)
```

## 
```{r}
apply(rand.mat, MARGIN = 2, FUN = sum)
apply(rand.mat, MARGIN = 1, FUN = var)
```

## `sapply()` 
- `sapply()` is used on `list`-objects.
```{r}
my.list <- list(A = c(4, 2, 1:3), B = "Hello", C = TRUE)
sapply(my.list, class)
```

## 
```{r}
my.list <- list(A = c(4, 2, 1:3), B = "Hello", C = TRUE)
sapply(my.list, range)
```
- It returns a vector or a matrix, depending on the output of the function
- Why is each element a character string?

##
- Any `data.frame` is also a `list`, where each column is one `list`-element.
```{r}
class(cats)
is.list(cats)
```

##
```{r}
print.default(cats[1:10, ])
```

##
- This means we can use `sapply` on data frames as well, which is often useful
```{r}
sapply(cats, class)
```
- Remember the for-loop to accomplish the same thing?


## `lapply()`
- `lapply()` is *exactly* the same as `sapply()`, but it returns a list instead of a vector
```{r}
lapply(cats, class)
```

# Writing your own functions

## What are functions?
- Functions are reusable pieces of code that take an input, do some computation on the input, and return an output
- We have already been using a lot of functions: code of the form `something()` is usually a function
```{r}
mean(1:6)
```

##
- We can make our own functions as follows
```{r}
squared <- function (x){
  xx <- x * x
  return(xx)
}

squared(4)
```

- The input `x` is called the (formal) *argument* of the function
- The output `xx` is called the *return value* of the function

##
- If there is no `return()` the last line is automatically returned, so (if we are lazy...) we can also just write

```{r}
squared <- function(x){
  x * x
}

squared(-2)
squared(1:3)
```

## {.smaller}

- We can use our own function with `apply()` just like all other functions
```{r}
rand.mat
apply(rand.mat, 2, squared)
```


## 
- What does the following function do? 
```{r}
areDivisible <- function(x, y){
  if (x > y){
    larger  <- x
    smaller <- y
  } else {
    larger  <- y
    smaller <- x
  }
  remainder <- larger %% smaller 
  remainder == 0
}
```

## {.build}
```{r}
areDivisible(3, 10)
areDivisible(3, 9)
areDivisible(9, 3)
```

## Default options
- Default options for some arguments are provided in many functions
- They allow us to provide an additional option, but if no choice is provided, we can choose standard behaviour
```{r}
areDivisible <- function(x, y, printInput = FALSE){
  if (printInput){
    cat("Testing if", x, "and", y, "are divisible: \n")
  }
  ((x %% y) == 0) | ((y %% x) == 0)
}
```
- Notice the sleek one-liner replacing the previous nine, but be carefull!

## {.smaller}

```{r}
areDivisible(3, 10)
areDivisible(3, 9, printInput = TRUE)
areDivisible(9, 3, printInput = FALSE)
```

# Practical

## Practical E

- Exercise 1 to 6 are about `for`-loops and conditional execution. These are important!
- Exerise 7 to 13 are about using the apply-family to summarise data in different ways. Exercises 7 and especially 8 are most important. Exercises 11-13 practise writing simple functions but includes a bit of string-handling.
- Exercise 14 is more challenging and should probably just be studied.